---
title: "Equitable Distribution Report"
author: "Sabrina xie"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

options(scipen = 10, digits = 2)

library(readxl)  #for excel files
library(tidyverse) #for data wrangling
library(ggplot2) #for figures
library(lubridate) #for date-times
library(tidycensus) #for mapping
library(tigris)
library(viridis) #for color maps

```

## Read in data

Data for median income, spending & collections from Eversource, UI in 2019, 2020 and 2021 filed under C&LM.

Data for demographics:

- Energy burden from EPA LEAD tool

- Justice40 disadvantaged communities from Climate and economic Justice Screening Tool

- Racial diversity from US Census Bureau 2020 Decennial Census

There are some problems with census tract numbers in the utility data so some demographic data does not collate properly.

```{r, include=FALSE, echo=FALSE}

loop <- c("2019","2020","2021")
all <- data.frame(
  `Census Tract`=character(0),
  Town=character(0),
  `Distressed Tract`=character(0),
  Size=character(0),
  Cat=character(0),
  collections=numeric(0),
  disbursements=numeric(0),
  Utility=character(0),
  Year=numeric(0),
  stringsAsFactors=FALSE
)

for(k in loop){
  
  #Paste in folder where files are located
  folder <- paste("C:/Users/xiesa/OneDrive - State of Connecticut/2019-2021 Equitable/",k,"/",sep="")
  
  #Run script to format files into 1 big data frame
  source("C:/Users/xiesa/OneDrive - State of Connecticut/2019-2021 Equitable/Read-data.R", local=knitr::knit_global())
  
  # the outputs of this script are data frames for each utility for large, small and HES data, titled "ES" or "UI"
  # it also creates "data" which is one large dataframe with ALL the collection and disbursement data for both utilities
  
  # the columns in "data" are: Census Tract, Town, Distressed Tract, Cat, collections, disbursements, Utility
  
  data <- data %>% 
    mutate(Year=as.numeric(k))
  
  all <- rbind(all,data)
  
}

data <- all

# Other metrics
source("C:/Users/xiesa/OneDrive - State of Connecticut/2019-2021 Equitable/Demo-data.R", local=knitr::knit_global())

data <- data %>%
  distinct() %>%  
  mutate(disbursements=replace_na(disbursements,0),collections=replace_na(collections,0))

race <- get_decennial(
  geography = "tract",
  state = "CT",
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = TRUE
) %>%
  mutate(percent = 100 * (value / summary_value)) %>% 
  filter(variable != "White") %>% 
  group_by(GEOID) %>% 
  mutate(percent = sum(percent)) %>% 
  ungroup() %>% 
  select(-variable,-value) %>% 
  distinct()

race2 <- race %>% select(GEOID,percent) %>% mutate(GEOID=as.numeric(GEOID)) %>% rename(`Census Tract`=GEOID)

data <- left_join(data,race2,by=c("Census Tract")) %>%  
  select(-geometry) %>% 
  rename(Racial.diversity2=percent,
         Racial.diversity3=Racial.diversity) %>% 
  mutate(Racial.diversity=ifelse(is.na(Racial.diversity2), Racial.diversity3,
                           ifelse(Racial.diversity2 >= 0, Racial.diversity2, NA))) %>% 
  select(-Racial.diversity2,-Racial.diversity3)

errors <- data[!complete.cases(data), ]
census <- read.csv("C:/Users/xiesa/OneDrive - State of Connecticut/2019-2021 Equitable/full list.csv") %>% rename(`Census Tract`=Census.Tract)
errors <- left_join(errors, census, by="Census Tract")

rm(list=setdiff(ls(), c("data","race")))

```

## Tables

### Distressed tracts

Overall trend toward closing gaps

- Distressed tracts: median income <60% state median income

- Small load: <100 kWh


```{r}

totals <- data %>% 
  filter(Size!="HES",
         Cat=="TOTAL") %>% 
  group_by(Utility,Year) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct()

tab <- data %>% 
  filter(Size!="HES", #large and small
         Cat=="TOTAL", #residential and C&I
         `Distressed Tract`=="YES") %>%  #distressed
  group_by(Utility,Year) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(collections=collections/totals$collections*100, #this is the proportion of total collections
         disbursements=disbursements/totals$disbursements*100, #this is the proportion of total disbursements
         equitable=case_when(collections>disbursements ~ "NO",
                             collections<disbursements ~ "YES")) #this is the comparison

print(tab)

tab<-pivot_longer(tab,cols=c("collections","disbursements"),names_to="Category",values_to="Pct")
  
  
print(ggplot(data=tab,aes(x=Year,y=Pct,color=Utility,linetype=Category))+
        geom_line()+
        geom_point(shape=4)+
        facet_grid(cols=vars(Utility)))


totals <- data %>% 
  filter(Size!="HES",
         Cat!="TOTAL") %>% 
  group_by(Utility,Year,Size,Cat) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct()

tab <- data %>% 
  filter(Size!="HES", #large and small
         Cat!="TOTAL", #residential and C&I
         `Distressed Tract`=="YES") %>%  #distressed
  group_by(Utility,Year,Size,Cat) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(collections=collections/totals$collections*100, #this is the proportion of total collections
         disbursements=disbursements/totals$disbursements*100, #this is the proportion of total disbursements
         equitable=case_when(collections>disbursements ~ "NO",
                             collections<disbursements ~ "YES")) #this is the comparison

print(tab)

tab<-pivot_longer(tab,cols=c("collections","disbursements"),names_to="Category",values_to="Pct")
  
  
print(ggplot(data=tab,aes(x=Year,y=Pct,color=Utility,linetype=Category))+
        geom_line()+
        geom_point(shape=4)+
        facet_grid(cols=vars(Size),rows=vars(Cat)))

```


### HES and HES-IE

It makes sense that distribution of HES-IE funds would be greater to distressed census tracts, but overall funds are not equitably distributed under HES program. Moderate distress appears to participate in HES-IE.

- HES-IE: Home Energy Solutions, income-eligible

- Moderately distressed tracts: median income 60-80% state median income

```{r}
totals <- data %>% 
  filter(Size=="HES") %>% 
  group_by(Utility,Year) %>% 
  mutate(collections=sum(collections)/2) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct()

tab <- data %>% 
  filter(Size=="HES",
         `Distressed Tract`=="YES") %>%  #distressed
  group_by(Utility,Year) %>% 
  mutate(collections=sum(collections)/2) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(collections=collections/totals$collections*100, #this is the proportion of total collections
         disbursements=disbursements/totals$disbursements*100, #this is the proportion of total disbursements
         equitable=case_when(collections>disbursements ~ "NO",
                             collections<disbursements ~ "YES")) #this is the comparison

print(tab)

tab<-pivot_longer(tab,cols=c("collections","disbursements"),names_to="Category",values_to="Pct")
  
  
print(ggplot(data=tab,aes(x=Year,y=Pct,color=Utility,linetype=Category))+
        geom_line()+
        geom_point(shape=4)+
        facet_grid(cols=vars(Utility))+
        labs(title="Distressed communities, HES & HES-IE total"))



totals <- data %>% 
  filter(Size=="HES") %>% 
  group_by(Utility,Year,Cat) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct()

tab <- data %>% 
  filter(Size=="HES",
         `Distressed Tract`=="YES") %>%  #distressed
  group_by(Utility,Year,Cat) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(collections=collections/totals$collections*100, #this is the proportion of total collections
         disbursements=disbursements/totals$disbursements*100, #this is the proportion of total disbursements
         equitable=case_when(collections>disbursements ~ "NO",
                             collections<disbursements ~ "YES")) #this is the comparison

print(tab)

tab<-pivot_longer(tab,cols=c("collections","disbursements"),names_to="Category",values_to="Pct")
  
  
print(ggplot(data=tab,aes(x=Year,y=Pct,color=Utility,linetype=Category))+
        geom_line()+
        geom_point(shape=4)+
        facet_grid(cols=vars(Utility),rows=vars(Cat))+
        labs(title="Distressed communities, HES & HES-IE"))



totals <- data %>% 
  filter(Size=="HES") %>% 
  group_by(Utility,Year,Cat) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct()

tab <- data %>% 
  filter(Size=="HES", 
         Moderate.distress=="YES") %>%  #moderately distressed
  group_by(Utility,Year,Cat) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(collections=collections/totals$collections*100, #this is the proportion of total collections
         disbursements=disbursements/totals$disbursements*100, #this is the proportion of total disbursements
         equitable=case_when(collections>disbursements ~ "NO",
                             collections<disbursements ~ "YES")) #this is the comparison

print(tab)

tab<-pivot_longer(tab,cols=c("collections","disbursements"),names_to="Category",values_to="Pct")
  
  
print(ggplot(data=tab,aes(x=Year,y=Pct,color=Utility,linetype=Category))+
        geom_line()+
        geom_point(shape=4)+
        facet_grid(cols=vars(Utility),rows=vars(Cat))+
        labs(title="Moderately distressed communities, HES & HES-IE"))

```


### Justice40 disadvantaged

Justice40 requires that 40% of funds flow to disadvantaged communities, which make up approximately 29% of the country. This means that these communities should receive 1.3x the average proportional investment.

- Disadvantaged: Communities that are marginalized, underserved, and overburdened by pollution. Burden thresholds include climate change, energy, health, housing, legacy pollution, transportation, water and wastewater, and workforce development.

- Black lines in the chart show targets based on 40% of budget and 1.3x proportional investment.

```{r}

totals <- data %>% 
  filter(Size!="HES",
         Cat=="TOTAL") %>% 
  group_by(Utility,Year) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct()

tab <- data %>% 
  filter(Size!="HES", #large and small
         Cat=="TOTAL", #residential and C&I
         `Justice40.disadvantaged`=="TRUE") %>%  #distressed
  group_by(Utility,Year) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(collections,disbursements,Utility,Year) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(collections=collections/totals$collections*100, #this is the proportion of total collections
         disbursements=disbursements/totals$disbursements*100, #this is the proportion of total disbursements
         equitable=case_when(collections>disbursements ~ "NO",
                             collections<disbursements ~ "YES")) #this is the comparison

print(tab)

tab<-pivot_longer(tab,cols=c("collections","disbursements"),names_to="Cat",values_to="Pct")
  
  
print(ggplot(data=tab,aes(x=Year,y=Pct,color=Utility,linetype=Cat))+
        geom_line()+
        geom_point(shape=4)+
        geom_hline(yintercept=40)+
        geom_line(data=tab[tab$Cat=="collections",],aes(x=Year,y=Pct*1.3),color="black")+
        facet_grid(cols=vars(Utility)))

```


## Maps & Baselines

```{r}

by.tract <- data %>% 
  filter(Cat != "total",
         Size != "HES") %>% 
  group_by(`Census Tract`,Year) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(-Size,-Cat) %>% 
  ungroup() %>% 
  distinct()

```

### Median household income

- Low: <60% of state median income
- Moderate: 60-80% of state median income
- High: >80% state median income

Must exceed baseline to be considered equitable.
Baseline is proportional investment based on portion of census tracts that fall under each category.

```{r, results='hide'}

med_income <- get_acs(
  geography = "tract", 
  variables = "B19013_001",
  state = "CT", 
  year = 2020,
  geometry = TRUE
)

ggplot(data = med_income, aes(fill = estimate/66270*100)) + 
  geom_sf(colour="white",lwd=0) + 
  scale_fill_viridis(option="viridis")+
  labs(title = "Median Household Income by Census Tract, 2020",
       caption = "Data source: 2020 1-year ACS, US Census Bureau",
       fill = "% state median income")+
  theme_minimal()+
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())

med_income <- med_income %>% 
  mutate(cat=case_when(
    estimate/66270<=0.6~"low",
    between(estimate/66270,0.6,0.8)~"moderate",
    estimate/66270>=0.8~"high"
  )) %>% 
  rename("Census Tract"="GEOID") %>%
  mutate(`Census Tract` = as.numeric(`Census Tract`))

ggplot(med_income) +
  geom_histogram(aes(x=estimate),bins=30)+
  geom_vline(xintercept=66270)+
  labs(caption="vertical line denotes state median income of $66,270")

med_income$cat <- factor(med_income$cat, levels = c("low","moderate","high"))

ggplot(med_income)+
  geom_bar(aes(x=cat),stat="count")

low.t <- length(which(med_income$cat=="low"))/length(med_income$cat)
mod.t <- length(which(med_income$cat=="moderate"))/length(med_income$cat)
high.t <- length(which(med_income$cat=="high"))/length(med_income$cat)


med_income.join <- left_join(by.tract,med_income,all.y=TRUE) %>% 
  group_by(Year,cat,Utility) %>% 
  mutate(collections=sum(collections)) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(Year,cat,Utility,collections,disbursements)%>% 
  distinct() %>% 
  filter(is.na(cat)==FALSE)

totals <- med_income.join %>% 
  group_by(Year,Utility) %>% 
  mutate(disbursements=sum(disbursements)) %>% 
  select(Year,Utility,disbursements) %>% 
  distinct() %>% 
  mutate(low=disbursements*low.t,
         moderate=disbursements*mod.t,
         high=disbursements*high.t) %>% 
  select(-disbursements)

totals<-pivot_longer(totals,cols=c("low","moderate","high"),names_to="cat",values_to="baseline")

med_income.join <- left_join(med_income.join,totals)

med_income.join$cat <- factor(med_income.join$cat, levels = c("low","moderate","high"))

ggplot(med_income.join, aes(y=disbursements,x=cat,fill=Utility))+
  geom_bar(stat="identity",position="dodge")+
  geom_errorbar(aes(ymin=baseline,ymax=baseline),width=1,position=position_dodge(0.9))+
  facet_grid(cols=vars(Year),rows=vars(Utility))+
  labs(caption="Black line denotes baseline")


  
```

### Racial diversity
```{r, results='hide'}

ggplot(data = race, aes(fill = percent)) + 
  geom_sf(colour="white",lwd=0) + 
  scale_fill_viridis(option="viridis")+
  labs(title = "Percent Population of Color by Census Tract, 2020",
       caption = "Data source: 2020 ACS, US Census Bureau",
       fill = "% POC estimate")+
  theme_minimal()+
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())

hist(race$percent)

```

### Energy burden

```{r, results='hide'}

LEAD <- read.csv("C:/Users/xiesa/OneDrive - State of Connecticut/2019-2021 Equitable/LEAD.csv") 
census_tracts <- tracts("CT",cb=TRUE,year="2010") %>% mutate(census_tract=as.numeric(paste(STATE,COUNTY,TRACT,sep="")))

LEAD <- LEAD %>% rename(census_tract=Geography.ID)
LEAD.joined <- left_join(census_tracts,LEAD,by="census_tract") %>% 
  filter( Avg..Energy.Burden....income. <100)

ggplot(data = LEAD.joined, aes(fill = Avg..Energy.Burden....income.)) + 
  geom_sf(colour="white",lwd=0) + 
  scale_fill_viridis(option="viridis")+
  labs(title = "Average Energy Burden by Census Tract, 2016",
       caption = "Data source: Low-Income Energy Affordability Data Tool",
       fill = "% of income")+
  theme_minimal()+
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())

```

### Disbursements & Collections

```{r fig.width=10,fig.height=10}

data.joined <- left_join(census_tracts,data,by=c("census_tract"="Census Tract")) %>% filter(Size!="HES",Cat=="TOTAL") %>% select(-Size) %>% group_by(census_tract,Year) %>% mutate(collections=sum(collections),disbursements=sum(disbursements)) %>% distinct()

b <- c(-Inf,1.35,Inf)
colors <- c('navyblue', 'orange', 'darkorange1')

data.joined<-pivot_longer(data.joined,cols=c("collections","disbursements"),names_to="Category",values_to="Value")

ggplot(data = data.joined, aes(fill = Value)) + 
  geom_sf(colour="white",lwd=0) + 
  scale_fill_viridis(option="viridis")+
  labs(title = "Proportion by Census Tract",
       caption = "Data source: C&LM program",
       fill = "$")+
  theme_minimal()+
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())+
  facet_grid(cols=vars(Category),rows=vars(Year))


```